Application-Specific Enhanced Morphological Analyzer \\for Arabic Text Mining

### 2011-02-26 tex/sarf.aclhlt.tex
presumably final version of Sarf

\begin{document}
\maketitle
\begin{abstract}
Natural language processing (NLP) uses morphological 
analysis to abstract and annotate text.
Often times, 
%      and in particular in the context of the Arabic language, 
annotations resulting from 
a branch of morphological analysis may not be
appropriate for the case study at hand. 
In this paper we present Sarf, an application-specific morphological analyzer for Arabic. 
Sarf allows an application-specific controller to control and refine
the state of the analysis on the fly. 
It uses non-deterministic finite state Automata
to analyze a stream of text. 
%Each machine corresponds to a valid analysis. 
Sarf uses concatenative analysis based on recursive affixes 
to better preserve part of speech information.
%We automated the analysis of three books of Islamic
%narrations (hadith) using Sarf, where we
%abstracted each book into a vector of complex structures.
Our results show better efficiency
compared to current analyzers.
\end{quote}
\end{abstract}

@@

%Automated analysis of Arabic data sets, including texts, 
%publications, records and digital media is essential
%with the huge digital Arabic content available nowadays. 
Morphological analysis is key in current automated 
analysis techniques for Arabic text. 
%Current morphological analyzers~\cite{Sughaiyer:04}
%use concatenative analysis when
%considering the internal structure 
%of an Arabic word and
%composing it into several {\em morphemes}. 
%A morpheme can be a {\em stem}, or an {\em affix}.
%An affix can be a {\em prefix, suffix, } or an {\em infix}.
%The analysis of one word may lead to several possible
%morphological solutions.
%\vocalize
%For instance, the word \RL{'a.hmadH}~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
%may have two valid morphological analyses. 
%The letter \RL{'a} may be a prefix and the word means 
%``I praise him'', or 
%%The letter \RL{'a} may also be 
%part of the stem \RL{'a.hmad} (a proper noun)
%and the word means ``his Ahmad''.
%
%\novocalize
The accuracy of the solutions suffer due to inherent difficulties
of morphological analysis of the Arabic language. 
For example, it is common practice to write Arabic text
without short vowels / diacritics. 
This greatly increases the ambiguity of Arabic text.
%(maybe reduce the couple of sentences below)
Moreover, Arabic letters can have up to 
four different forms
corresponding to their position in a word, i.e, beginning,
middle, end of word and separate forms. 
This allows the phrase 
\noTrnoVocRL{il_A\nospace almdrsT}
to be visually recognizable
as two separate words \noVocRL{il_A} (to) and \noVocRL{almdrsT} (the school) 
without the need of a space in between. 
The reason is the first word \noVocRL{il_A} ends with
\noVocRL{_A} a non-connecting letter. 
These words,
referred to as ``run-on'' words~\cite{Buckwalter:04},
occur often, and greatly increase the
difficulty of tokenization.

%{\bf Sarf.~~~}
In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
has been submitted with the paper and will be available online as an open source tool.},
 a {\em novel application-specific %efficient 
enhanced %(instead of 'efficient' to engulf both efficiency and other refinements)
morphological analyzer} that uses 
non-deterministic finite state Automata 
driven by an application-specific controller.
Each machine in Sarf takes one letter at a time as input
and represents a valid morphological analysis.
The application-specific controller prunes false positives
early in the run by making a decision at every input letter
to control the machines.
%The controller we present in Section~\ref{sec:controller}
%is an FSA. However, as we explain in Section~\ref{sec:ndfsa},
%it can be a logical entity of any type.
The controller can be an FSA or any other %logical entity 
programming entity of any type,
since Automatas may be a very complex solution for some simple applications.
  %(maybe re-word this sentence)

%Each alive machine in Sarf takes as input one letter at a time 
%from a text stream.
In addition, to being application-specific, Sarf has introduced 
many enhancements to the field of arabic morphological analysis. 
For example,
Sarf does not assume that the word at hand is a token and
performs tokenization on the fly based on morphological correctness.
Up to our knowledge, this is the first morphological analyzer that 
handles the ``run-on'' words problem.

Sarf uses concatenative analysis with a novel %(maybe remove novel bc authors did not feel it is much)
{\em recursive affixes} refinement,
where an affix can be a sequence of other affixes which are concatenatively compatible.
%An atomic affix $a$ is a standalone affix
%that is not the concatenation of other affixes. 
%We define an affix $f: a | a f$ to be either
%an atomic affix $a$ or an atomic affix concatenated
%with an affix. 
Sarf introduces recursive affixes in order to
better retain part of speech information and enhance the 
efficiency of affix matching.

In addition, Sarf supports disambiguation using diacritics when they are available.
This feature, to the best of our knowledge is the first attempt to benefit from 
diacritics instead of ignoring them. Prior to Sarf, the trend was to preprocess 
the text to remove any such `unwelcomed' letter. %Nonetheless, Sarf regards them as a richness to be utilized that should not be disregarded.

%Moreover, since Sarf performs tokenization on the fly, Sarf is not just capable of solving
%the problem of ``run-on'' words, but also of handling Multiword Expressions (MWE). 
Moreover, since Sarf does not perform white-space delimited tokenization, Sarf is capable of 
handling Multiword Expressions (MWEs).
MWEs are defined in~\cite{MWE}
``as idiosyncratic interpretations
that cross word boundaries (or spaces)''. 
%By this term, we refer to frequent cases when a group of words form together a single
%item with a semantical meaning. In this paper, we focus on the example of compound names such as 
%\noVocRL{`bid alkarym}. 
%In current analyzers, this phrase is analyzed word by word and in best cases is 
%considered to contain two names, instead of being understood to be a single but compound name.

Our focus when building Sarf, was not on improving the semantical rules and 
expanding the morphological alternatives, instead we worked on adding support 
for higher level control by the specific application utilizing the lower level 
morphological analyzer. %In other words, the application-specific controller, has 
% full control over and directs the morphological analysis of Sarf. 
% (sentence should be reworded differently, but is needed to explain why we mention 
% higher and lower accuracy hypothesis, which seems vague otherwise)
%
%We have noted that many NLP case studies need the 
%morphological analyzer to answer queries that do not need 
%high accuracy at the morphological analysis level.
%The analysis at the higher levels 
%(i.e. answering the application-specific query)
%can often compensate for 
%tolerable inaccuracy at lower morphological levels. 
%For example, if a query concerns proper names and the 
%analyzer is considering a 
%prefix that connects only to a verb,
%%the analyzer can answer without considering the 
%%rest of the word.
%we do not need to further analyze the rest of the word 
%to provide a negative answer.
%
%We find evidence to our hypothesis in~\cite{Maamouri:10}.
%%The addition of a new corpus to the Arabic Tree Bank 
%%(ATB)~\cite{Maamouri:04}
%%%required compliance with the DARPA GALE program 
%%%including several NLP tasks such as English translation 
%%%annotations and word-level alignment of Arabic and English. 
%%%The new data 
%%challenged the existing processing and 
%%annotation guidelines. 
%%This led to refining 
%%SAMA~\cite{Kulick:10} with 
%%alternative guidelines. 
%%This refinement was achieved only after the NLP 
%%task interacted with the morphological analyzer,
%%a flexibility our analyzer is designed to provide for the application-specific controller on the run. %(check if this sentence makes sense)
%It reports that a refined version of SAMA~\cite{Kulick:10} was achieved when NLP task interacted with the morphological analyzer.
%%a flexibility our analyzer is designed to provide for the application-specific controller on the run.
%In addition, 
%the work of Habash~\shortcite{Habash:06} reports that different 
%types of morphological analyses behave differently on the same %case study. 
%data set.
%The analyzers may report different solutions or 
%may present the solutions in different orders. 
%Clearly, some of 
%the presented solutions and their order of presentation may not be 
%appropriate for the %case study.
%application needs, which is an issue that Sarf tackles. % (maybe reworded)
We strongly believe that by being
an application-specific morphological analyzer that allows an application-specific 
controller to intervene at every decision to guide, use, order, and 
refine the morphological analysis, Sarf is of high utility.

%We validate our hypothesis using a case study from the Islamic 
%literature, introduced later in Section~\ref{sec:results}.
In addition to all its enhanced analysis methods, Sarf still
outperformed current morphological analyzers in terms of running time.
These results will be presented in Section~\ref{sec:results}.

We make the following key contributions: 
\begin{enumerate}
\item {\bf application-specific analysis:}  We designed and implemented
a application-specific morphological analyzer where a controller
guides the analysis. 
%\item {\bf Exhaustive analysis:} Since we have an application-specific 
%controller that can decide on the fly whether an analysis is 
%accepted or not, we can afford to keep all valid analyses modulo
%those pruned by the case controller. 
%We do that using non-deterministic FSAs. 
\item {\bf Recursive affixes:}
We refine the concatenative analysis of 
Buckwalter~\shortcite{Buckwalter:02} and define affixes
recursively. 
This allows Sarf to perform better and 
maintain compositional part of speech (POS) tags.
\item {\bf Diacritic-based Disambiguation:}
Diacritics are used to disambiguate morphological
alternatives using any existant diacritics.
\end{enumerate}

@@

The survey in~\cite{Sughaiyer:04} compares
several morphological analyzers. 
Analyzers such as~\cite{Khoja:01,Darwish:02} 
target specific applications in the 
analyzer itself or use a specific set of POS tags
as their reference.
Sarf differs in that it is a general morphological 
analyzer that reports all possible solutions. 
It is application-specific in the sense that a controller 
prunes these solutions.

Current morphological analyzers such as %(maybe disambiguators also added here according to reviewer comments)
Buckwalter~\shortcite{Buckwalter:02},
Beesley~\shortcite{Beesley:01},
SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} 
take as input white space delimited tokens~\cite{Kulick:10},
consider them as words,
and enumerate all possible solutions. 
%This approach has several problems. 
However, we note the following points that deserve consideration:
%(We need not to mention especially point 2 as a "problem" according to reviewer comments)
First, a white space delimited token may have 
more than one word or just a part of a MWE.
%Second, the exhaustive enumeration may hurt performance and may
%not be necessary or appropriate
%in some case studies as noted in~\cite{Maamouri:10}. 
Second, in many applications, the exhaustive enumeration may not be needed or appropriate as noted in~\cite{Maamouri:10}; consequently, adding unnecessary running time.
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
% (reviewer wants to distinguish MADA as a tool for morphological disambiguation instead of analysis)
use machine learning and support vector machines (SVM) 
to enhance the accuracy of the morphological analysis at the expense 
of efficiency.
Xerox rules can be compiled into specialized finite state
machine (FSM) based analyzers as described in~\cite{Beesley:03}.
However, the efficiency of the resulting analyzer depends on the
way the Xerox rules are written. 
This requires deep knowledge and insight from the user
in compilation techniques, context free grammars, 
and morphological analysis.
%context free grammar rules that can be automatically compiled
%into efficient non-deterministic finite state Automata (NDFSA).

Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
It differs from Buckwalter in that it defines a shorter list of affixes
and a longer list of affix compatibility rules to allow 
affixes to be defined recursively
so that we can have prefix-prefix an suffix-suffix 
concatenations.
This allows Sarf to better maintain and propagate 
the POS tags associated with affixes. 
%This information is highly 
%redundant with Buckwalter.% and that may lead to consistency issues. 
Sarf keeps all the possible analyses alive where each analysis
corresponds to a set of decisions and positions in the text. 
Buckwalter produces a set of solutions for each token 
and the several analysis threads of the text need to be 
generated as a product of the token solutions.

Similar to Sarf, both Beesley~\shortcite{Beesley:01} and Attia~\shortcite{Attia:05} use
finite state Automata (FSA). 
Beesley reports that the number of Automata generated by a compiler
for Xerox rules can not be controlled and also reports a difficulty 
to compose the FSAs into a single framework. 
Sarf however constructs an NDFSA from
the controller and the prefix, stem, and suffix deterministic 
transducers to solve the problem.
Sarf also differs in that it uses recursive affixes in the 
affix transducers, and in that the transducers consider
input from the controller to modify their behavior. 
%With Sarf it is easy to control the FSA with a application-specific
%controller. 
Beesley requires recompilation of the Xerox rules. 
Attia has researched extensively MWEs handling methods~\cite{Attia:06} and 
automated extending their dictionary coverage~\cite{Attia:10}. 
% (I am not sure how clear this statement is)

Similar to The MADA+TOKAN toolkit, Sarf thinks of
the several valid morphological analyses as a richness that 
should be exploited at a higher abstract level rather than
an inaccuracy that should be corrected. 
Sarf differs in that it provides an interface for the 
application to interfere in judging the analyses much earlier and
at a finer granularity level. 
In Sarf, there is no need for a separate tokenizer such as
TOKAN since each solution keeps a stack of positions
that partition text into morphemes.
%The application-specific controller can build the trace 
%when needed.

@@

Our approach is close to the refinement of SAMA~\cite{Maamouri:10}.
SAMA was refined to interact with
the ATB~\cite{Maamouri:04} project after the addition of a large 
new corpus. 
%We find supporting evidence in this decision to our hypothesis
%about the utility of a application-specific analyzer. 
The algorithmic changes in SAMA were
done manually and worked in integration with the ATB format. 
Sarf is an application-specific controller approach that can modify 
the behavior of the analyzer on the fly and can interact
with any application as needed. 
In Sarf, the analyzer remains general, and the 
application-specific controller specializes the analysis.

The AMIRA~\cite{Diab:07} analyzer uses 
a language independent SVM based analysis. 
The SVM analysis learns features from the input text 
and does not make use of any information 
in the target query.

Like ElixirFM~\cite{Otakar:07}, Sarf builds on the lexicon
of the Buckwalter analyzer. 
Sarf also uses deterministic parsing with tries 
to implement the affix and stem transducers. 
We think that the inferential-realizational approach 
of ElixirFM
that is highly compatible with the Arabic linguistic 
description~\cite{Badawi:04}
can benefit from many features unique to the Arabic language.
Sarf leaves implementing that to the application-specific controller
since in many cases the abstraction needs only a partial 
linguistic model of the Arabic language. 
%We can compare the application-specific controller in Sarf to a high level
%component in the ElixirFM domain specific language.

@@

Sarf extends the lexicon of Buckwalter~\shortcite{Buckwalter:02} with
proper names and location names extracted from different online 
sources~\footnote{\href{http://alasmaa.net/}{http://alasmaa.net/ }, 
\href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
as well as biblical sources~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.

Sarf encodes the prefix, suffix and stem lexicons into
three deterministic transducers and 
uses a non-deterministic construction
of the three to compute all valid morphological analyses.
We first present a running example of Sarf.
The running example explains how Sarf works,
implements recursive affixes, and
handles ``run-on'' words.
%Then we discuss how Sarf builds and composes the different FSAs.

\subsection{Sarf example}
\label{sec:example}

@@

%{\bf Sarf example.}

The diagram in Figure~\ref{f:example}
illustrates the operation of $\Phi$, the Sarf FSA,
when parsing the
string \noTrnoVocRL{wsyl`b-h-aa\nospace al-lA`bwn}.
%\RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
\footnote{ 
\noVocRL{n}%
\noVocRL{w}
\noVocRL{b}
\noVocRL{`}
\noVocRL{a} 
\noVocRL{l}
\noVocRL{l}
\noVocRL{a} %
\noVocRL{a}
\noVocRL{h}
\noVocRL{b}
\noVocRL{`}
\noVocRL{l}
\noVocRL{y}
\noVocRL{s}
\noVocRL{w}
in separate form to ease following the example
}

Boxes and circles denote accept and regular
states respectively.
The edges are transitions with labels corresponding to
the input letters.
The reject states are omitted for clarity and are denoted
by the absence of corresponding transitions.

Subfigures (a), (b), and (c) in Figure~\ref{f:example}
represent ${\cal P}$ the prefix ,
${\cal S}$ the stem, and ${\cal X}$ the suffix
transducers respectively. 
The symbol $\epsilon$ represents an empty string and is 
the source of non-determinism.

We start with $\Phi$ at the root square state in 
${\cal P}$ which is an accept state. 
The $\epsilon$-edge connects ${\cal P}$
to ${\cal S}$ and transitions from any accept state
in ${\cal P}$ to the root state in ${\cal S}$.
The $\epsilon$-edge that connects ${\cal S}$ to ${\cal X}$
follows the same behavior.

When there are two valid transitions such as \noTrnoVocRL{w} 
and $\epsilon$ in the start case, 
$\Phi$ spawns an exact copy of itself $\Psi$. 
$\Phi$ makes one transition, and $\Psi$ makes the other. 
Each of $\Phi$ and $\Psi$ represent a valid analysis so far. 
An FSA ($\Phi$ or $\Psi$) dies when it reaches a reject state.
In our example, if there were no stems that start 
with the letter \noTrnoVocRL{w}, $\Psi$ will die. 
In reality, $\Psi$ will die when the input 
is at \noTrnoVocRL{wsyl`}.

Note that the affix transducers allow recursive 
affixes. 
For example, \noTrnoVocRL{w} will result in an accept state
that transitions to ${\cal S}$.
When \noTrnoVocRL{s} follows, we move to another accept state in 
${\cal P}$ corresponding to the prefix \noTrnoVocRL{ws}. 
The same applies to ${\cal X}$.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Natural language processing (NLP) uses morphological 
analysis to annotate text.
Often times, 
%      and in particular in the context of the Arabic language, 
annotations resulting from 
a branch of morphological analysis may not be
appropriate for the NLP application. 
We present Sarf, 
%an application-specific morphological analyzer for Arabic. 
an Arabic application-specific morphological analyzer, which
%Sarf
allows a controller to refine
the analysis on the fly. 
It uses non-deterministic finite state Automata % (NDFSA)
to analyze a stream of text. 
%Each machine corresponds to a valid analysis. 
Sarf uses concatenative analysis 
%based on
with  
recursive affixes 
to better preserve part of speech (POS) information.
It solves the issue of ``run-on'' words 
and utilizes diacritics for disambiguation.
%We automated the analysis of three books of Islamic
%narrations (hadith) using Sarf, where we
%abstracted each book into a vector of complex structures.
Our results show better efficiency
compared to current analyzers.
\end{quote}
\end{abstract}

@@

Morphological analysis is key in current automated 
analysis techniques for Arabic text. 
The accuracy of the solutions suffers due to inherent difficulties
of morphological analysis of Arabic. 
For example, it is common practice to write Arabic text
without short vowels (diacritics) 
which greatly increases ambiguity.
Moreover, tokenization is not always 
trivial.
For example,
\noTrnoVocRL{il_A\nospace almdrsT}
can be visually recognizable
as two separate words \noVocRL{il_A} (to) and \noVocRL{almdrsT} (the school) 
without the need of a space in between. 
The reason is that
\noTrnoVocRL{_A} is a non-connecting letter. 
Such 
``run-on'' words~\cite{Buckwalter:04},
occur often.

%{\bf Sarf.~~~}
In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
has been submitted with the paper and will be available 
as an open source tool.}, 
a {\em novel application-specific efficient 
%(maybe 'enhanced' instead of 'efficient' to engulf both efficiency and other refinements)
morphological analyzer} that uses 
non-deterministic finite state Automata (NDFSA)
driven by an application-specific controller.
Each machine in Sarf takes one letter at a time as input
and represents a valid morphological analysis.
%The application-specific controller prunes false positives
The controller prunes false positives
early in the run by making a decision at every input letter
to control the machines.
%The controller we present in Section~\ref{sec:controller}
%is an FSA. However, as we explain in Section~\ref{sec:ndfsa},
%it can be a logical entity of any type.
%since Automatas may be a %very complex
%non-necessary solution for some simple applications.
%(maybe re-word this sentence)
%depending on application needs.

%(check if position of this paragraph seems in correct place)
%Our focus when building Sarf, was not on improving the semantical rules and 
%expanding the morphological alternatives, instead we worked on adding support 
%for higher level control by the specific application utilizing the lower level 
%morphological analyzer. %In other words, the application-specific controller, has 
% full control over and directs the morphological analysis of Sarf. 
% (sentence should be reworded differently, but is needed to explain why we mention 
% higher and lower accuracy hypothesis, which seems vague otherwise)
%
%We have noted that many NLP case studies need the 
%morphological analyzer to answer queries that do not need 
%high accuracy at the morphological analysis level.
%The analysis at the higher levels 
%(i.e. answering the application-specific query)
%can often compensate for 
%tolerable inaccuracy at lower morphological levels. 
%For example, if a query concerns proper names and the 
%analyzer is considering a 
%prefix that connects only to a verb,
%%the analyzer can answer without considering the 
%%rest of the word.
%we do not need to further analyze the rest of the word 
%to provide a negative answer.
%
%We find evidence to our hypothesis in~\cite{Maamouri:10}.
%%The addition of a new corpus to the Arabic Tree Bank 
%%(ATB)~\cite{Maamouri:04}
%%%required compliance with the DARPA GALE program 
%%%including several NLP tasks such as English translation 
%%%annotations and word-level alignment of Arabic and English. 
%%%The new data 
%%challenged the existing processing and 
%%annotation guidelines. 
%%This led to refining 
%%SAMA~\cite{Kulick:10} with 
%%alternative guidelines. 
%%This refinement was achieved only after the NLP 
%%task interacted with the morphological analyzer,
%%a flexibility our analyzer is designed to provide for the application-specific controller on the run. %(check if this sentence makes sense)
%It reports that a refined version of SAMA~\cite{Kulick:10} was achieved when NLP task interacted with the morphological analyzer.
%%a flexibility our analyzer is designed to provide for the application-specific controller on the run.
%In addition, 
%the work of Habash~\shortcite{Habash:06} reports that different 
%types of morphological analyses behave differently on the same %case study. 
%data set.
%The analyzers may report different solutions or 
%may present the solutions in different orders. 
%Clearly, some of 
%the presented solutions and their order of presentation may not be 
%appropriate for the %case study.
%application needs, which is an issue that Sarf tackles. % (maybe reworded)
We hypothesize that in many NLP applications, the morphological analyzer may
benefit from the characteristics of the query at hand, 
to greatly enhance the analysis.
Consequently, %we believe that by 
%being 
%allowing
an application-specific %morphological 
%analyzer that allows %an application-specific 
%the
controller that intervenes at every decision to guide, use, order, and 
refine the morphological analysis is of high utility.

%Each alive machine in Sarf takes as input one letter at a time 
%from a text stream.
%Apart from being application-specific, 
Sarf %has
also  
introduces 
several enhancements. %to the field of Arabic morphological analysis. 
%the following enhancements:
%For example,
First,
Sarf %does not assume that the word at hand is a token and
performs tokenization on the fly based on morphological correctness.
Up to our knowledge, this is the first %morphological 
analyzer that 
handles %the ``run-on'' words problem. 
``run-on'' words. 
%
%Moreover, since Sarf performs tokenization on the fly, Sarf is not just capable of solving
%the problem of ``run-on'' words, but also of handling Multiword Expressions (MWE). 
On the fly tokenization also handles
Multiword Expressions (MWEs)~\cite{MWE} such as \noVocRL{`bid alkarym}.
%MWEs are defined in~\cite{MWE} as
%which are 
%``idiosyncratic interpretations
%that cross word boundaries (or spaces)''. 
%By this term, we refer to frequent cases when a group of words form together a single
%item with a semantical meaning. In this paper, we focus on the example of compound names such as 
%\noVocRL{`bid alkarym}. 
%In current analyzers, this phrase is analyzed word by word and in best cases is 
%considered to contain two names, instead of being understood to be a single but compound name.

Second, 
Sarf uses concatenative analysis with a novel %(maybe remove novel bc authors did not feel it is much)
{\em recursive affixes} refinement,
%where an affix can be a sequence of other affixes which are concatenatively compatible.
where an affix can be a sequence of other compatible affixes.
%An atomic affix $a$ is a standalone affix
%that is not the concatenation of other affixes. 
%We define an affix $f: a | a f$ to be either
%an atomic affix $a$ or an atomic affix concatenated
%with an affix. 
%Sarf introduces recursive affixes in order to
%This refinement is introduced to better retain
Hence, it better retains
%part of speech information
POS information  
and enhances the 
efficiency of affix matching. 
Third, Sarf supports disambiguation using 
%diacritics when they are available.
available diacritics, instead of the traditional approach that ignores them.
%This feature, to the best of our knowledge is the first attempt to benefit from 
%To the best of our knowledge,
%Up to our knowledge, this is the first attempt to benefit from
%diacritics instead of ignoring them. 
%Nonetheless, Sarf regards them as a richness to be utilized that should not be disregarded.

%In addition to all its enhanced analysis methods, Sarf still
%Sarf also
%outperformed current morphological analyzers in terms of running time,
%%These results will be presented 
%as shown in Section~\ref{sec:results}.
Sarf is also more efficient than current morphological analyzers, 
as shown in Section~\ref{sec:results}.

%%We make the following 
%We summarize our key contributions: 
%\begin{enumerate}
%\item {\bf application-specific analysis:}  We designed and implemented
%a application-specific morphological analyzer where a controller
%guides the analysis. 
%%\item {\bf Exhaustive analysis:} Since we have an application-specific 
%%controller that can decide on the fly whether an analysis is 
%%accepted or not, we can afford to keep all valid analyses modulo
%%those pruned by the case controller. 
%%We do that using non-deterministic FSAs. 
%\item {\bf Recursive affixes:}
%We refine the concatenative analysis of 
%Buckwalter~\shortcite{Buckwalter:02} and define affixes
%recursively. 
%This allows Sarf to perform better and 
%maintain compositional part of speech (POS) tags.
%\item {\bf Diacritic-based Disambiguation:}
%Diacritics are used to disambiguate morphological
%alternatives using any existant diacritics.
%\end{enumerate}

@@

The survey in~\cite{Sughaiyer:04} compares
several morphological analyzers. 
Analyzers such as~\cite{Khoja:01,Darwish:02} 
target specific applications 
%in the analyzer itself or use a specific set of POS tags as their reference
or limit themselves to a specific set of POS tags. 
Sarf differs in that it is a general morphological 
analyzer that reports all possible solutions. 
It is application-specific in the sense that a controller 
prunes these solutions.

Buckwalter~\shortcite{Buckwalter:02},
Beesley~\shortcite{Beesley:01},
SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} 
consider white-space delimited tokens~\cite{Kulick:10}
as words.
%This suffers has several problems. 
%However, we note the following: %(I thought of this change not to include point 2 as a "problem" according to reviewer comments)
This fails when a white-space delimited token has
more than one word or is part of an MWE.
%Second, the exhaustive enumeration may hurt performance and may
%not be necessary or appropriate
%in some case studies as noted in~\cite{Maamouri:10}. 
These analyzers also exhaustively enumerate all 
possible morphological solutions 
which may not be appropriate in many applications
~\cite{Maamouri:10}. 
%consequently, adding unnecessary running time.

Others such as 
AMIRA~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
use machine learning and support vector machines (SVM) 
to disambiguate the morphological analysis at the expense 
of efficiency.
Xerox rules can be compiled into specialized 
finite state Automata (FSA)~\cite{Beesley:03}.
%FSA-based analyzers~\cite{Beesley:03}.
The efficiency of the resulting analyzer depends on the
way the rules are written. 
This requires deep knowledge and insight from the user
in compilation techniques, context free grammars, 
and morphological analysis.
%context free grammar rules that can be automatically compiled
%into efficient non-deterministic finite state Automata (NDFSA). 
Beesley reports that the number of Automata generated by a compiler
for Xerox rules can not be controlled and also reports a difficulty 
to compose the FSAs into a single framework. 
%Beesley requires recompilation of the Xerox rules. 
% (I am not sure if it is ok to have the above removed)
Sarf however constructs an NDFSA from 
%the controller and 
the prefix, stem, and suffix deterministic 
transducers to solve the problem.
Sarf also differs in that 
%it uses recursive affixes in the affix transducers, and in that 
the transducers consider
input from the controller to modify their behavior.
The user can express the controller,not necessarily an FSA, in a programming language such as C++.

Sarf builds upon the lexicon of Buckwalter. %\shortcite{Buckwalter:02}, %(already cited above)
%It differs from Buckwalter in that it defines a shorter list of affixes
It defines a shorter list of affixes
and a longer list of 
%affix 
compatibility rules to define 
affixes recursively and allow 
prefix-prefix and suffix-suffix 
concatenations.
%This allows Sarf to better maintain and propagate 
This better maintains and propagates 
the tags associated with affixes. 
%This information is highly 
%redundant with Buckwalter.% and that may lead to consistency issues. 
% (check if necessary to keep the 2 sentences below)
%Sarf keeps all the possible analyses alive where each analysis
%corresponds to a set of decisions and positions in the text. 
%Buckwalter produces a set of solutions for each token 
%and the several analysis threads of the text need to be 
%generated as a product of the token solutions. 
Similar to Sarf, Attia uses FSAs and handles more and qualitatively different MWEs
~\cite{Attia:06,Attia:10}.
%Attia has researched extensively MWEs handling methods~\cite{Attia:06} and 
%automated extending their dictionary coverage~\cite{Attia:10}. 
% (I am not sure how clear this statement is)

%Similar to %the 
%MADA+TOKAN, Sarf uses
%the several valid morphological analyses as 
%%a richness that should 
%richness at a higher abstract level, and does not view them as 
%inaccuracies to be corrected. 
Sarf differs from MADA+TOKAN in that it provides an interface for the 
application to interfere in judging the analyses much earlier and
at a finer granularity level. 
%In Sarf, there is no need for a separate tokenizer such as
Sarf does not need a separate tokenizer such as
TOKAN since each solution keeps a stack of positions
that partition text into morphemes.
%The application-specific controller can build the trace 
%when needed.

@@

SAMA~\cite{Maamouri:10} was refined after the addition of a new corpus to 
the ATB~\cite{Maamouri:04}, since the new data challenged the existing processing and 
annotation guidelines.
The modification involved interaction between SAMA and the NLP 
task and resulted in alternative guidelines. 
Sarf supports such an interaction on the fly.

%The AMIRA %~\cite{Diab:07}
%analyzer uses 
%a language independent SVM-based analysis. 
%%The SVM 
%This 
%analysis learns features from the input text 
%but does not make use of any information 
%in the target query.

Like ElixirFM~\cite{Otakar:07}, Sarf builds on the lexicon
of Buckwalter. %analyzer. 
Sarf also uses deterministic parsing with tries 
to implement the affix and stem transducers. 
The inferential-realizational approach 
of ElixirFM
is highly compatible with the Arabic linguistic 
description~\cite{Badawi:04}.
%can benefit from many features unique to the Arabic language.
Sarf leaves implementing that to the %application-specific 
controller
since in many applications the abstraction needs only a partial 
linguistic model. 
%We can compare the application-specific controller in Sarf to a high level
%component in the ElixirFM domain specific language.

@@

%Sarf extends the lexicon of Buckwalter~\shortcite{Buckwalter:02} with
%proper names and location names extracted from different online 
%sources~\footnote{\href{http://alasmaa.net/}{http://alasmaa.net/ }, 
%\href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
%as well as biblical sources~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.

Sarf encodes the prefix, suffix and stem lexicons into
three deterministic transducers and 
uses a non-deterministic construction
of the three to compute all valid morphological analyses.
We first present a running example 
%of Sarf. The running example 
that 
explains how Sarf works,
implements recursive affixes, and
handles ``run-on'' words.
%Then we discuss how Sarf builds and composes the different FSAs.

%\subsection{Sarf example}
%\label{sec:example}

@@

{\bf Sarf example.~~~}
The diagram in Figure~\ref{f:example}
illustrates the operation of $\Phi$, the Sarf FSA,
when parsing the
string \noTrnoVocRL{wsyl`b-h-aa\nospace al--lA`bwn} \noArRL{wsyl`b-h-aa al--lA`bwn}.
%\RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
%\footnote{ 
%%\noVocRL{n}%
%%\noVocRL{w}
%%\noVocRL{b}
%%\noVocRL{`}
%%\noVocRL{a} 
%%\noVocRL{l}
%%\noVocRL{l}
%%\noVocRL{a} %
%%\noVocRL{a}
%%\noVocRL{h}
%%\noVocRL{b}
%%\noVocRL{`}
%%\noVocRL{l}
%%\noVocRL{y}
%%\noVocRL{s}
%%\noVocRL{w}
%\noTrnoVocRL{n}%
%\noTrnoVocRL{w}
%\noTrnoVocRL{b}
%\noTrnoVocRL{`}
%\noTrnoVocRL{a} 
%\noTrnoVocRL{l}
%\noTrnoVocRL{l}
%\noTrnoVocRL{a} %
%\noTrnoVocRL{a}
%\noTrnoVocRL{h}
%\noTrnoVocRL{b}
%\noTrnoVocRL{`}
%\noTrnoVocRL{l}
%\noTrnoVocRL{y}
%\noTrnoVocRL{s}
%\noTrnoVocRL{w}
%in separate form% to ease following the example
%}

Boxes and circles denote accept and regular
states respectively.
The edges are transitions with labels corresponding to
the input letters.
The reject states are omitted for clarity and are denoted
by the absence of corresponding transitions. 
Subfigures (a), (b), and (c) in Figure~\ref{f:example}
represent ${\cal P}$ the prefix ,
${\cal S}$ the stem, and ${\cal X}$ the suffix
transducers respectively. 
The symbol $\epsilon$ represents an empty string and is 
the source of non-determinism.

We start with $\Phi$ at the root accept state 
in ${\cal P}$. 
%The $\epsilon$-edge connects ${\cal P}$
%to ${\cal S}$ and transitions from any accept state
%in ${\cal P}$ to the root state in ${\cal S}$.
%The $\epsilon$-edge that connects ${\cal S}$ to ${\cal X}$
%follows the same behavior. 
The $\epsilon$-edges connect ${\cal P}$
to ${\cal S}$  and ${\cal S}$ to ${\cal X}$.
They transition from any accept state
in the current transducer to the root state of the target transducer.

When there are two valid transitions such as \noTrnoVocRL{w} 
and $\epsilon$ in the start case, 
$\Phi$ spawns an exact copy of itself $\Psi$. 
$\Phi$ makes one transition, and $\Psi$ makes the other. 
Each of $\Phi$ and $\Psi$ represent a valid analysis so far. 
An FSA %($\Phi$ or $\Psi$) 
dies when it reaches a reject state.
In our example, if there were no stems that start 
with the letter \noTrnoVocRL{w}, $\Psi$ will die. 
In reality, $\Psi$ will die when the input 
is at \noTrnoVocRL{wsyl`}.
